\huge The human visual system preserves the hierarchy of 2-dimensional pattern regularity

### 2021-03-11 Writeup/kohler_clarke_2021.tex
minor spelling mistakes corrected

The two datasets we present here make it possible to assess the extent to which both behavior and brain responses follow the hierachy of complexity expressed by the subgroup relationships. Based on previous brain imaging work showing that patterns with more axes of symmetry produce greater activity in visual cortex \citep{RN1170, RN1959, RN1725, keefe_2018}, we hypothesized that more complex groups would produce larger SSVEPs. For the psychophysical data, we hypothesized that more complex groups would lead to shorter symmetry detection thresholds, based on previous data showing that under a fixed presentation time, discriminability increases with the number of symmetry axes in the pattern \citep{wagemans_1991}. Our results confirm both hypotheses, and show that activity in human visual cortex is remarkably consistent with the hierarchical relationships between the wallpaper groups, with SSVEP amplitudes and psychophysical thresholds following these relationships at a level that is far beyond chance. The human visual system thus appears to encode all of the fundamental symmetries using a representational structure that closely approximates the subgroup relationships from group theory.

@@

The stimuli used in our two experiments were generated from random-noise textures, which made it possible to generate multiple exemplars from each of the wallpaper groups, as described in detail elsewhere \citep{RN1725}. We generated control stimuli matched to each exemplar in the main stimulus set, by scrambling the phase but maintaining the power spectrum. All wallpaper groups are inherently periodic because of their repeating lattice structure. Phase scrambling maintains this periodicity so the phase-scrambled control images all belong to group \textit{P1} regardless of group membership of the original exemplar. \textit{P1} contains no symmetries other than translation, while all other groups contain translation in combination with one or more of the other three fundamental symmetries (reflection, rotation, glide reflection) \citep{RN1425}. In our SSVEP experiment, this stimulus set allowed us to isolate brain activity specific to the symmetry structure in the exemplar images from activity associated with modulation of low-level features, by alternating exemplar images and control exemplars. In this design, responses to structural features beyond the shared power spectrum, including any symmetries other than translation, are isolated in the odd harmonics of the image update frequency \citep{RN1725,RN1540, RN1954}. Thus, the combined magnitude of the odd harmonic response components can be used as a measure of the overall strength of the visual cortex response.

@@

We now wanted to test our two hypotheses about how SSVEP amplitudes and threshold durations would follow subgroup relationships, and thereby quantify the degree to which our two measurements were consistent with the group theoretical hierarchy of complexity. We tested each hypothesis using the same approach. We first fitted a Bayesian model with wallpaper group as a factor and participant as a random effect. We fit the model separately for SSVEP RMS and psychophysical data, and then computed posterior distributions for the difference between supergroup and subgroup. These difference distributions allowed us to compute the conditional probability that the supergroup would produce (a) larger RMS and (b) a shorter threshold durations, when compared to the subgroup. The posterior distributions are shown in Figure 2 for the SSVEP data, and in Figure 3 for the psychophysical data, which distributions color-coded according to conditional probability. For both data sets our hypothesis is confirmed: For the overwhelming majority of the 63 subgroup relationships, supergroups are more likely to produce larger symmetry-specific SSVEPs and shorter symmetry detection threshold durations, and in most cases the conditional probability of this happening is extremely high.

We also ran a control analysis using (1) odd-harmonic SSVEP data from a six-electrode ROI over parietal cortex (see Supplementary Figure 1.1) and (2) even-harmonic SSVEP data from the same occipital ROI that was used in our primary analysis. By comparing these two control analysis to our primary SSVEP analysis, we can address the specify of our effects in terms of location (occipital cortex vs parietal cortex) and harmonic (odd vs even). For both control analyses (plotted in Supplementary Figures 3.3 and 3.4), the correspondence between data and subgroup relationships was substantially weaker than in the primary analysis. We can quantify the strength of the assocation between the data and the subgroup relationships, by asking what proportion of subgroup relationships that reach or exceed a range of probability thresholds. This is plotted in Figure 4, for our psychophysical data, our primary SSVEP analysis and our two control SSVEP analyses. It shows that odd-harmonic SSVEP data from the occipital ROI and symmetry detection threshold durations both have a strong association with the subgroup relationships such that a clear majority of the subgroups survive even at the highest threshold we consider ($p(\Delta>0|data)>0.99$). The assocation is far weaker for the two control analyses.

SSVEP data from four of the wallpaper groups (\textit{P2}, \textit{P3}, \textit{P4} and \textit{P6}) was previously published as part of our earlier demonstration of parametric responses to rotation symmetry in wallpaper groups\citep{RN1725}. We replicate that result using our Bayesian approach, and find an analogous parametric effect in the psychophysical data (see Supplementary Figure 4.1). We also conducted an analysis testing for an effect of index in our two datasets, and found that subgroup relationships with higher indices tended to produce greater pairwise differences between the subgroup and supergroup, for both SSVEP RMS and symmetry detection thresholds (see Supplementary Figure 4.2). The effect of index is relatively weak, but the fact that there is a measurable index effect can nonetheless be taken as preliminary evidence that representations of symmetries in wallpaper groups may be compositional.

Finally, we conducted a correlation analysis comparing SSVEP and psychophysical data, and found a reliable correlation ($R^2 = 0.44$, Bayesian confidence interval [0.28, 0.55]). The correlation reflects an inverse relationship: For subgroup relationships where the supergroup produces a much \textit{larger} SSVEP amplitude than the subgroup, the supergroup also tends to produce a much \textit{smaller} symmetry detection threshold. This is consistent with our hypotheses about how the two measurements relate to symmetry representations in the brain, and suggests that our brain imaging and psychophysical measurements are at least to some extent tapping into the same underlying mechanisms.

@@

We observe a reliable correlation between our brain imaging and psychophysical data. This suggests that the two measurements reflect the same underlying symmetry representations in visual cortex. While it should be noted that the correlation is relatively modest ($R^2 = 0.44$), we note that this may be partly due to the fact that the same individuals did not participate in the two experiments. Future work in which behavioral and brain imaging data are collected from the same participants, will help further establish the connection between the two measurements, and tease apart any additional complexity that may not have been captured by the summary statistics we applied here. It has recently been demonstrated that \textit{W}, a measure of perceptual goodness derived from a holographic model of regularity \citep{van_der_helm_1996}, can predict EEG responses \citep{makin_2016} and perceptual discrimination performance \citep{nucci_2007} for patterns that contain symmetry and other types of regularity. The model was formulated based on dot patterns with symmetry axes centered on a single spatial location. It will important to determine if and how \textit{W} can computed for our random-noise based wallpaper textures where combinations of symmetries tile the plane, and test how well it can explain behavioral and brain responses to wallpapers.

@@

The 17 wallpaper groups are completely regular, and relatively rare in the visual environment, especially when considering distortions due to perspective and occlusion. Near-regular textures, however, abound in the visual world, and can be approximated as deformed versions of the wallpaper groups \citep{RN1519}. The correspondence between visual cortex responses and group theory demonstrated here may indicate that the visual system represents visual textures using a similar scheme, with the wallpaper groups serving as anchor points in representational space. This framework resembles norm-based encoding strategies that have been proposed for other stimulus classes, most notably faces \citep{RN435}, and leads to the prediction that adaptation to wallpaper patterns should distort perception of near-regular textures, similar to the aftereffects found for faces \citep{RN1768}. Field biologists have demonstrated that animals respond more strongly to exaggerated versions of a learned stimulus, referred to as “supernormal” stimuli \citep{RN1775}. In the norm-based encoding framework, wallpaper groups can be considered \textit{supertextures}, exaggerated examples of the near-regular textures that surround us. Artists may consciously or unconsciously create supernormal stimuli, to capture the essence of the subject and evoke strong responses in the audience \citep{RN1764}. Wallpaper groups are visually compelling and symmetries have been widely used in human artistic expression going back to the Neolithic age \citep{RN1949}. If wallpapers are in fact supertextures, this prevalence may be a direct result of the strategy the human visual system has adopted for texture encoding.

@@

\subsection*{Stimulus Generation}
Exemplars from the different wallpaper groups were generated using a modified version of the methodology developed by Clarke and colleagues\citep{RN172} that we have described in detail elsewhere\citep{RN1725}. Briefly, exemplar patterns for each group were generated from random-noise textures, which were then repeated and transformed to cover the plane, according to the symmetry axes and geometric lattice specific to each group. The use of noise textures as the starting point for stimulus generation allowed the creation of an almost infinite number of distinct exemplars of each wallpaper group. To make individual exemplars as similar as possible we replaced the power spectrum of each exemplar with the median across exemplars within a group. We then generated control exemplars that had the same power spectrum as the exemplar images by randomizing the phase of each exemplar image. The phase scrambling eliminates rotation, reflection and glide-reflection symmetries within each exemplar, but the phase-scrambled images inherent the spectral periodicity arising from the periodic tiling. This means that all control exemplars, regardless of which wallpaper group they are derived from, are transformed into another symmetry group, namely \textit{P1}. \textit{P1} is the simplest of the wallpaper groups, and contains only translations of a region whose shape derives from the lattice. Because the different wallpaper groups have different lattices, \textit{P1} controls matched to different groups have different power spectra. Our experimental design takes these differences into account by comparing the neural responses evoked by each wallpaper group to responses evoked by the matched control exemplars.

@@

\subsection*{Psychophysics Procedure}
The experiment consisted of 16 blocks, one for each of the wallpaper groups (excluding \textit{P1}). We used a two-interval forced choice approach. In each trial, participants were presented with two stimuli (one of which was the wallpaper group for the current block of trials, the other being \textit{P1}), one after the other (inter-stimulus interval of 700ms). After each stimuli had been presented, it was masked with white noise for 300ms. After both stimuli had been presented, participants made a response on the keyboard to indicate whether they thought the first or second image contained more symmetry. Each block started with 10 practice trials, (stimulus display duration of 500ms) to allow participants to familiarise themselves with the current block's wallpaper pattern. If they achieved an accuracy of 9/10 in these trials they progressed to the rest of the block, otherwise they carried out another set of 10 practise trials. This process was repeated until the required accuracy of 9/10 was obtained. The rest of the block consisted of four interleaved staircases (using the QUEST algorithm \citep{watson1983quest}, full details given in the SI) of 30 trials each. On average, a block of trials took around 10 minutes to complete.

@@

Bayesian analysis was carried out using R (v3.6.1) \citep{R} with the \texttt{brms} package (v2.9.0) \citep{burkner2017} and rStan (v2.19.2 \citep{rStan}). The data from each experiment were modelled using a Bayesian generalised mixed effect model with wallpaper group being treated as a 16 level factor, and random effects for participant. The SSVEP data and symmetry detection threshold durations were modelled using log-normal distributions with weakly informative, $ \mathcal{N}(0, 2)$, priors. After fitting the model to the data, samples were drawn from the posterior distribution of the two datasets, for each wallpaper group. These samples were then recombined to calculate the distribution of differences for each of the 63 pairs of subgroup and supergroup. These distributions were then summarised by computing the conditional probability of obtaining a positive (negative) difference, $p(\Delta | \text{data})$. For further technical details, please see the Supplementary Materials at \texttt{https://osf.io/f3ex8/} where the full R code, model specification, prior and posterior predicitive checks, and model diagnoastics, can be found.

>>>>

The two datasets we present here make it possible to assess the extent to which both behavior and brain responses follow the hierarchy of complexity expressed by the subgroup relationships. Based on previous brain imaging work showing that patterns with more axes of symmetry produce greater activity in visual cortex \citep{RN1170, RN1959, RN1725, keefe_2018}, we hypothesized that more complex groups would produce larger SSVEPs. For the psychophysical data, we hypothesized that more complex groups would lead to shorter symmetry detection thresholds, based on previous data showing that under a fixed presentation time, discriminability increases with the number of symmetry axes in the pattern \citep{wagemans_1991}. Our results confirm both hypotheses, and show that activity in human visual cortex is remarkably consistent with the hierarchical relationships between the wallpaper groups, with SSVEP amplitudes and psychophysical thresholds following these relationships at a level that is far beyond chance. The human visual system thus appears to encode all of the fundamental symmetries using a representational structure that closely approximates the subgroup relationships from group theory.

@@

The stimuli used in our two experiments were generated from random-noise textures, which made it possible to generate multiple exemplars from each of the wallpaper groups, as described in detail elsewhere \citep{RN1725}. We generated control stimuli matched to each exemplar in the main stimulus set, by scrambling the phase but maintaining the power spectrum. All wallpaper groups are inherently periodic because of their repeating lattice structure. Phase scrambling maintains this periodicity, so the phase-scrambled control images all belong to group \textit{P1} regardless of group membership of the original exemplar. \textit{P1} contains no symmetries other than translation, while all other groups contain translation in combination with one or more of the other three fundamental symmetries (reflection, rotation, glide reflection) \citep{RN1425}. In our SSVEP experiment, this stimulus set allowed us to isolate brain activity specific to the symmetry structure in the exemplar images from activity associated with modulation of low-level features, by alternating exemplar images and control exemplars. In this design, responses to structural features beyond the shared power spectrum, including any symmetries other than translation, are isolated in the odd harmonics of the image update frequency \citep{RN1725,RN1540, RN1954}. Thus, the combined magnitude of the odd harmonic response components can be used as a measure of the overall strength of the visual cortex response.

@@

We now wanted to test our two hypotheses about how SSVEP amplitudes and threshold durations would follow subgroup relationships, and thereby quantify the degree to which our two measurements were consistent with the group theoretical hierarchy of complexity. We tested each hypothesis using the same approach. We first fitted a Bayesian model with wallpaper group as a factor and participant as a random effect. We fit the model separately for SSVEP RMS and psychophysical data and then computed posterior distributions for the difference between supergroup and subgroup. These difference distributions allowed us to compute the conditional probability that the supergroup would produce (a) larger RMS and (b) a shorter threshold durations, when compared to the subgroup. The posterior distributions are shown in Figure 2 for the SSVEP data, and in Figure 3 for the psychophysical data, which distributions color-coded according to conditional probability. For both data sets our hypothesis is confirmed: For the overwhelming majority of the 63 subgroup relationships, supergroups are more likely to produce larger symmetry-specific SSVEPs and shorter symmetry detection threshold durations, and in most cases the conditional probability of this happening is extremely high.

We also ran a control analysis using (1) odd-harmonic SSVEP data from a six-electrode ROI over parietal cortex (see Supplementary Figure 1.1) and (2) even-harmonic SSVEP data from the same occipital ROI that was used in our primary analysis. By comparing these two control analysis to our primary SSVEP analysis, we can address the specify of our effects in terms of location (occipital cortex vs parietal cortex) and harmonic (odd vs even). For both control analyses (plotted in Supplementary Figures 3.3 and 3.4), the correspondence between data and subgroup relationships was substantially weaker than in the primary analysis. We can quantify the strength of the association between the data and the subgroup relationships, by asking what proportion of subgroup relationships that reach or exceed a range of probability thresholds. This is plotted in Figure 4, for our psychophysical data, our primary SSVEP analysis and our two control SSVEP analyses. It shows that odd-harmonic SSVEP data from the occipital ROI and symmetry detection threshold durations both have a strong association with the subgroup relationships such that a clear majority of the subgroups survive even at the highest threshold we consider ($p(\Delta>0|data)>0.99$). The association is far weaker for the two control analyses.

SSVEP data from four of the wallpaper groups (\textit{P2}, \textit{P3}, \textit{P4} and \textit{P6}) was previously published as part of our earlier demonstration of parametric responses to rotation symmetry in wallpaper groups\citep{RN1725}. We replicate that result using our Bayesian approach, and find an analogous parametric effect in the psychophysical data (see Supplementary Figure 4.1). We also conducted an analysis testing for an effect of index in our two datasets and found that subgroup relationships with higher indices tended to produce greater pairwise differences between the subgroup and supergroup, for both SSVEP RMS and symmetry detection thresholds (see Supplementary Figure 4.2). The effect of index is relatively weak, but the fact that there is a measurable index effect can nonetheless be taken as preliminary evidence that representations of symmetries in wallpaper groups may be compositional.

Finally, we conducted a correlation analysis comparing SSVEP and psychophysical data and found a reliable correlation ($R^2 = 0.44$, Bayesian confidence interval [0.28, 0.55]). The correlation reflects an inverse relationship: For subgroup relationships where the supergroup produces a much \textit{larger} SSVEP amplitude than the subgroup, the supergroup also tends to produce a much \textit{smaller} symmetry detection threshold. This is consistent with our hypotheses about how the two measurements relate to symmetry representations in the brain, and suggests that our brain imaging and psychophysical measurements are at least to some extent tapping into the same underlying mechanisms.

@@

We observe a reliable correlation between our brain imaging and psychophysical data. This suggests that the two measurements reflect the same underlying symmetry representations in visual cortex. While it should be noted that the correlation is relatively modest ($R^2 = 0.44$), we note that this may be partly due to the fact that the same individuals did not participate in the two experiments. Future work in which behavioral and brain imaging data are collected from the same participants, will help further establish the connection between the two measurements, and tease apart any additional complexity that may not have been captured by the summary statistics we applied here. It has recently been demonstrated that \textit{W}, a measure of perceptual goodness derived from a holographic model of regularity \citep{van_der_helm_1996}, can predict EEG responses \citep{makin_2016} and perceptual discrimination performance \citep{nucci_2007} for patterns that contain symmetry and other types of regularity. The model was formulated based on dot patterns with symmetry axes centered on a single spatial location. It will be important to determine if and how \textit{W} can computed for our random-noise based wallpaper textures where combinations of symmetries tile the plane, and test how well it can explain behavioral and brain responses to wallpapers.

@@

The 17 wallpaper groups are completely regular, and relatively rare in the visual environment, especially when considering distortions due to perspective and occlusion. Near-regular textures, however, abound in the visual world, and can be approximated as deformed versions of the wallpaper groups \citep{RN1519}. The correspondence between visual cortex responses and group theory demonstrated here may indicate that the visual system represents visual textures using a similar scheme, with the wallpaper groups serving as anchor points in representational space. This framework resembles norm-based encoding strategies that have been proposed for other stimulus classes, most notably faces \citep{RN435}, and leads to the prediction that adaptation to wallpaper patterns should distort perception of near-regular textures, similar to the aftereffects found for faces \citep{RN1768}. Field biologists have demonstrated that animals respond more strongly to exaggerated versions of a learned stimulus, referred to as “supernormal” stimuli \citep{RN1775}. In the norm-based encoding framework, wallpaper groups can be considered \textit{supertextures}, exaggerated examples of the near-regular textures that surround us. Artists may consciously or unconsciously create supernormal stimuli, to capture the essence of the subject and evoke strong responses in the audience \citep{RN1764}. Wallpaper groups are visually compelling, and symmetries have been widely used in human artistic expression going back to the Neolithic age \citep{RN1949}. If wallpapers are in fact supertextures, this prevalence may be a direct result of the strategy the human visual system has adopted for texture encoding.

@@

\subsection*{Stimulus Generation}
Exemplars from the different wallpaper groups were generated using a modified version of the methodology developed by Clarke and colleagues\citep{RN172} that we have described in detail elsewhere\citep{RN1725}. Briefly, exemplar patterns for each group were generated from random-noise textures, which were then repeated and transformed to cover the plane, according to the symmetry axes and geometric lattice specific to each group. The use of noise textures as the starting point for stimulus generation allowed the creation of an almost infinite number of distinct exemplars of each wallpaper group. To make individual exemplars as similar as possible we replaced the power spectrum of each exemplar with the median across exemplars within a group. We then generated control exemplars that had the same power spectrum as the exemplar images by randomizing the phase of each exemplar image. The phase scrambling eliminates rotation, reflection and glide-reflection symmetries within each exemplar, but the phase-scrambled images inherent the spectral periodicity arising from the periodic tiling. This means that all control exemplars, regardless of which wallpaper group they are derived from, are transformed into another symmetry group, namely \textit{P1}. \textit{P1} is the simplest of the wallpaper groups and contains only translations of a region whose shape derives from the lattice. Because the different wallpaper groups have different lattices, \textit{P1} controls matched to different groups have different power spectra. Our experimental design takes these differences into account by comparing the neural responses evoked by each wallpaper group to responses evoked by the matched control exemplars.

@@

\subsection*{Psychophysics Procedure}
The experiment consisted of 16 blocks, one for each of the wallpaper groups (excluding \textit{P1}). We used a two-interval forced choice approach. In each trial, participants were presented with two stimuli (one of which was the wallpaper group for the current block of trials, the other being \textit{P1}), one after the other (inter-stimulus interval of 700ms). After each stimulus had been presented, it was masked with white noise for 300ms. After both stimuli had been presented, participants made a response on the keyboard to indicate whether they thought the first or second image contained more symmetry. Each block started with 10 practice trials, (stimulus display duration of 500ms) to allow participants to familiarise themselves with the current block's wallpaper pattern. If they achieved an accuracy of 9/10 in these trials they progressed to the rest of the block, otherwise they carried out another set of 10 practise trials. This process was repeated until the required accuracy of 9/10 was obtained. The rest of the block consisted of four interleaved staircases (using the QUEST algorithm \citep{watson1983quest}, full details given in the SI) of 30 trials each. On average, a block of trials took around 10 minutes to complete.

@@

Bayesian analysis was carried out using R (v3.6.1) \citep{R} with the \texttt{brms} package (v2.9.0) \citep{burkner2017} and rStan (v2.19.2 \citep{rStan}). The data from each experiment were modelled using a Bayesian generalised mixed effect model with wallpaper group being treated as a 16-level factor, and random effects for participant. The SSVEP data and symmetry detection threshold durations were modelled using log-normal distributions with weakly informative, $ \mathcal{N}(0, 2)$, priors. After fitting the model to the data, samples were drawn from the posterior distribution of the two datasets, for each wallpaper group. These samples were then recombined to calculate the distribution of differences for each of the 63 pairs of subgroup and supergroup. These distributions were then summarised by computing the conditional probability of obtaining a positive (negative) difference, $p(\Delta | \text{data})$. For further technical details, please see the Supplementary Materials at \texttt{https://osf.io/f3ex8/} where the full R code, model specification, prior and posterior predictive checks, and model diagnostics, can be found.
